Modelling BCG vaccination in the UK

### 2019-12-11 rmds/01-introduction.Rmd
minor corrections complete
dodd minor corrections up to 6

Tuberculosis (TB) is one of the oldest human diseases, with recorded cases in ancient Egypt, renaissance Europe, and in the modern day across the globe. It is thought that roughly one-third of the world’s population has been infected with TB, with 1% of the world’s population being infected annually. However, the vast majority of these cases will never develop active disease. This reservoir of disease presents a challenge for control and eradication as, even if transmission can be halted, new cases will still occur for many years to come. While many people might consider TB to be a problem of the past in England, in 2017 there were 5,900 notified cases, the majority of which occurred in vulnerable communities; where incidence rates can be as much as 15 times higher than in the general population.[@PHE2017] Globally TB remains the leading cause of death from infectious disease.[@TheWorldHealthOrganization:2018va]

@@

The remainder of this chapter outlines: the theoretical framework used in this thesis; the aims and objects that were used to motivate this thesis; the chapter structure of this thesis; and the output from this thesis.

@@

Mechanistic mathematical models provide an assumption based framework for understanding the transmission of infectious diseases.[@Anderson1991] Mechanistic models can be used to simplify complex real-world systems, whilst retaining a linkage to real-world processes.[@Keeling2007] They are unlike statistical models, which instead focus on modelling the underlying structure of the data generally without reference to the real-world processes. There are multiple mechanistic modelling approaches, the most common being compartmental based models and individual based models.[@Anderson1991; @Keeling2007] Both of these approaches can be represented as  deterministic, semi-stochastic, or fully stochastic - both of the latter include a random component whilst the former does not. Recently, mechanistic models have been combined with statistical models to account for the fact that events may be only partially observed.[@King2016] Mechanistic models have an advantage over statistical models in that they can more easily be used to compare alternative scenarios over long periods of time for which observed data does not exist.

In this thesis, a partially observed semi-stochastic compartmental model of TB is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals at a rate dictated by the concentration of susceptibles in the population. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

* **Chapter \@ref(model-fitting):** In this chapter the model developed in the previous chapter is fitted to the routine surveillance data (Chapter \@ref(data)) using bayesian methods. Multiple scenarios are considered.

@@

* Abbott S., Christensen H., Brooks-Pollock E. *Reassessing the evidence for universal school-age Bacillus Calmette Guerin (BCG) vaccination in England and Wales*, doi: https://doi.org/10.1101/567511

@@

* **Assessing the Evidence for Universal Bacillus Calmette-Guèrin (BCG) Vaccination in England** - Research and Applied Epidemiology Scientific Conference 2016, Warwick, United Kingdom. Received best abstract from an early career researcher. Link: https://www.samabbott.co.uk/talk/phe-applied_epi-2016/

>>>>

Tuberculosis (TB) is one of the oldest human diseases, with recorded cases in ancient Egypt, renaissance Europe, and in the modern day across the globe.[@Stone2009] It is thought that roughly one-third of the world’s population has been infected with TB, with 1% of the world’s population being infected annually. However, the vast majority of these cases will never develop active disease. This reservoir of disease presents a challenge for control and eradication as, even if transmission can be halted, new cases will still occur for many years to come. While many people might consider TB to be a problem of the past in England, in 2017 there were 5,900 notified cases, the majority of which occurred in vulnerable populations; where incidence rates can be as much as 15 times higher than in the general population.[@PHE2017] Globally, TB remains the leading cause of death from infectious disease.[@TheWorldHealthOrganization:2018va]

@@

The remainder of this chapter outlines: the theoretical framework used in this thesis; the aims and objectives that were used to motivate this thesis; the chapter structure of this thesis; and the output from this thesis.

@@

Mechanistic mathematical models provide an assumption based framework for understanding the transmission of infectious diseases.[@Anderson1991] Mechanistic models can be used to simplify complex real-world systems, whilst retaining a linkage to real-world processes.[@Keeling2007] They are unlike statistical models, which instead focus on modelling the underlying structure of the data generally without reference to the real-world processes. There are multiple mechanistic modelling approaches, the most common being compartmental based models and individual based models.[@Anderson1991; @Keeling2007] Both of these approaches can be represented as  deterministic, semi-stochastic (deterministic with some stochastic elements),[@Funk2016a] or fully stochastic - both of the latter include a random component whilst the former does not. Recently, mechanistic models have been combined with statistical models to account for the fact that events may be only partially observed.[@King2016] Mechanistic models have an advantage over statistical models in that they can more easily be used to compare alternative scenarios over long periods of time for which observed data does not exist.

In this thesis, a partially observed semi-stochastic compartmental model of TB is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals at a rate dictated by the concentration of susceptibles in the population. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King2016] for implementation details using R.

@@

* **Chapter \@ref(model-fitting):** In this chapter the model developed in the previous chapter is fitted to the routine surveillance data (Chapter \@ref(data)) using Bayesian methods. Multiple scenarios are considered.

@@

* Abbott S., Christensen H., Brooks-Pollock E. *Reassessing the evidence for universal school-age BCG vaccination in England and Wales*, doi: https://doi.org/10.1101/567511

@@

* **Assessing the Evidence for Universal BCG Vaccination in England** - Research and Applied Epidemiology Scientific Conference 2016, Warwick, United Kingdom. Received best abstract from an early career researcher. Link: https://www.samabbott.co.uk/talk/phe-applied_epi-2016/

### 2019-12-11 rmds/02-background.Rmd
dodd minor corrections up to 6

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active (symptomatic) TB in their lifetime.[@Houben:2016jp] Of those with active TB around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, TB risk factors, TB treatment, global impact, and the impact TB in England and Wales are explored.

@@

TB is primarily a respiratory disease (pulmonary TB) caused by the bacterium Mycobacterium TB, although it can also affect other parts of the body (extra-pulmonary TB). TB spreads via airborne droplets that are expelled when individuals with active pulmonary TB cough. After infection with TB, 5-10% of individuals develop primary disease within 1-2 years of exposure. Children are more likely to develop active disease and to develop it more quickly than adults.[@TheWorldHealthOrganization:2018va] The remaining 90-95% of individuals then enter a latent stage in which they passively carry TB mycobacterium. Reactivation of bacilli can then occur many years later due to a loss of immune control.[@Gideon2011a]

@@

```{r global-regional-tb-summary, fig.cap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017. Globally incidence rates have been declining since the early 2000s but this decline varies with region.", fig.scap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, viridis_end = 1) + theme(text = element_text(size = 12))
```

@@

TB remains one of the top 10 causes death worldwide, leading to 1.7 million deaths in 2016 alone.[@TheWorldHealthOrganization:2018va]  Whilst the absolute number of deaths due to TB has fallen since 2000, the average annual global rate of decline in TB mortality rates was only 2.9% between 2000-2016. However, unlike the trend observed for incidence rates, the year-on-year decline of TB mortality rates has remained consistent in all regions (Figure \@ref(fig:global-regional-mortality-summary)). Several regions, including Africa and Europe, have seen TB mortality rates fall to below 50% of those in 2000.

@@

There is an ongoing global co-epidemic of HIV and TB, with people living with HIV accounting for 1.4 million TB cases in 2016. 22% of deaths from TB were in those living with HIV. Whilst this is a global problem, it is a particular issue in sub-Saharan Africa with over 60% (95% CI: 55%-64%) of incidence TB cases in South Africa also having HIV (Figure \@ref(fig:map-global-tb-with-hiv)). This compares to a global mean of 9.1% (95% CI: 6.0%-13.0%) and a mean of 26.7% (95% CI: 17.4%-38.1%) in Africa.

@@

Multi-drug-resistant TB (MDR-TB), which is defined as being resistant to at least isoniazid and rifampin, made up 4.6% of all new TB cases globally in 2015 (480,000). It can be acquired both through treatment failure and through transmission. Treatment requires the use of second line antibiotics, which often have more severe side effects and are more likely to fail, with only 52% successfully treated globally compared to 83% for drug susceptible TB.[@WHOTB2016] As for HIV co-infection, drug resistance is globally heterogeneous with some regions, like countries in the former USSR, having a much higher proportion of drug resistant cases. Figure \@ref(fig:map-global-tb-with-rr) shows the country level proportion of cases with at least rifampicin resistance and highlights the higher level of rifampicin resistance in countries formerly in the USSR. 92% of rifampicin cases in Russia in 2018 also had MDR-TB, which is comparable to the global median of 81% (Table \@ref(tab:prop-rr-with-mdr-tab)). Across all regions, Europe had the highest median percentage of rifampicin cases with MDR-TB (87%), with the Western Pacific having the lowest (78%).

@@

```{r prop-rr-with-mdr-tab, warning = FALSE}
summarise_tb_burden(countries = "Russia", 
                    metric = "e_mdr_pct_rr_new", verbose = FALSE,
                    year = 2018, stat = "median") %>% 
  mutate_at(.vars = c("e_mdr_pct_rr_new", "e_mdr_pct_rr_new_lo", "e_mdr_pct_rr_new_hi"),
            funs(ifelse(. > 100, 100, .))) %>% 
  arrange(desc(area)) %>% 
  mutate(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)` = 
           pretty_ci(e_mdr_pct_rr_new,
                     e_mdr_pct_rr_new_lo, 
                     e_mdr_pct_rr_new_hi, digits = 1)) %>% 
  mutate(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)` = 
           ifelse(area == "Russian Federation", paste(pretty_round(e_mdr_pct_rr_new), "***"),
                  as.character(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)`)
                  )) %>% 
  select(Area = area, `Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)`) %>% 
  knitr::kable(
               caption = "Percentage (\\%) of rifampicin resistant TB cases that have multi-drug resistant TB in Russia and regional medians, with interquartile ranges.",
        booktabs = TRUE) %>% 
  add_footnote(label = c("** Multi-Drug Resistant TB", "*** Only point estimates available"), notation = "none") %>% 
  kable_styling(font_size = 8, latex_options = "HOLD_position") %>% 
  {.}
```

@@

The Bacillus Calmette–Guérin (BCG) vaccine was first given to humans in 1921 and remains the only licensed vaccine for TB.[@Medicine2013] The BCG vaccine is a live vaccine and was developed by weakening a strain of Mycobacterium bovis, which is commonly found in cows, over a period of 13 years.[@TheWorldHealthOrganization:2018va] Serious side effects are rare, although a small scar at the injection site is common. Initially, public acceptance was slow, with low take up until after the Second World War. However, controversy remains and several countries have scaled back, or retargeted, their usage in recent years. This section details the action, effectiveness, duration of protection, effects and usage of the BCG vaccine.

@@

The BCG primarily acts by directly preventing the development of active, symptomatic disease. However, there is some evidence to suggest that the BCG vaccine also provides partial protection against initial infection.[@Roy2014] There is no evidence that BCG vaccination post infection with TB provides protection from developing active TB disease.[@TheWorldHealthOrganization:2018va]

@@

A meta-analysis of RCTs indicated that increased protection is associated with distance from the equator.[@Mangtani2014a] One hypothesis for this is that there is a greater density of mycobacterium near the equator that may mask, or block, the protection offered by the BCG vaccine.[@Zwerling2011] Recently it has been found that much of this latitude effect may be due to stringency in tuberculin skin testing (TST), with lower stringency near the equator.[@TheWorldHealthOrganization:2018va] TST screening tests for the presence of TB infection but may give a false positive if the subject has been exposed to other mycobacteria or the BCG vaccine. Reduced stringency would lead to a greater number of TB positive individuals being vaccinated. These individuals would then receive no protection from the vaccine and would lead to a reduced estimate of the effectiveness of the vaccine overall. Regardless of the mechanism this effect means that vaccination early in life maximises the protection conferred by the BCG vaccine in countries with evidence of decreasing protection with age.

@@

In addition to its effect on TB outcomes, the BCG vaccine has also been found to be effective at preventing leprosy (with a risk ratio (RR_ of 0.45 (95% CI: 0.34-0.56)), with some evidence that this protection was stronger in those vaccinated before 15 years of age.[@TheWorldHealthOrganization:2018va] Additionally, there is some evidence that the BCG vaccine can provide protection against Non-TB mycobacteria (NTM) infections, with an estimated effectiveness of 50%.[@TheWorldHealthOrganization:2018va] This protection has been shown to greatly reduce the incidence of NTM lymph node inflammation and protect against Buruli ulcers for the first 12 months following vaccination (RR, 0.50 [95% CI, .37–.69]). A study also found that individuals with Buruli ulcers are less likely to develop osteomyelitis (bone infection) if they have a BCG scar (RR, 0.36 (95% CI: 0.22 to 0.58)).[@Zimmermann:2018io]

@@

The BCG vaccine is one of the mostly widely-used vaccines worldwide, with approximately 100m doses given annually.[@The2004] However, due to the variable estimates of BCG efficacy, vaccination has been controversial since its development. The WHO recommends vaccination for all neonates as early as possible after birth in high burden settings. Vaccination in low burden settings is dependent on the country specific epidemiology of TB.[@WHO2017; @TheWorldHealthOrganization:2018va] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014] Historically, different strategies have been utilized worldwide. These include universal vaccination of those at most risk of onwards transmission and high-risk group vaccination targeting either neonates or children.[@Abubakar2013]

@@

As of 2011, among 180 countries with available data, 157 countries recommended universal BCG vaccination. The remaining 23 countries had either never implemented a universal programme or have switched to targeted vaccination of high risk individuals.[@Zwerling2011a] Most countries began universal programmes between the 1940s and 1980s due to high levels of TB incidence and strong evidence of the effectiveness of the BCG vaccine.[@Hart1972] In the last 20 years 49 of these countries reported changing their vaccination programme with 27 countries reporting major changes in the last 10 years.[@Zwerling2011a] Globally, in countries that have BCG vaccination policies in place, coverage is estimated to be between 70% and 100%.

>>>>

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active (symptomatic) TB in their lifetime.[@Houben:2016jp] Of those with active TB around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of active cases occur in less economically developed countries and may not be diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, TB risk factors, TB treatment, global impact, and the impact TB in England and Wales are explored.

@@

TB is primarily a respiratory disease (pulmonary TB) caused by the bacterium *Mycobacterium tuberculosis*, although it can also affect other parts of the body (extra-pulmonary TB). TB spreads via airborne droplets that are expelled when individuals with active pulmonary TB cough. After infection with TB, 5-10% of individuals develop primary disease within 1-2 years of exposure. Children are more likely to develop active disease and to develop it more quickly than adults.[@TheWorldHealthOrganization:2018va] The remaining 90-95% of individuals then enter a latent stage in which they passively carry TB mycobacterium. Reactivation of bacilli can then occur many years later due to a loss of immune control.[@Gideon2011a]

@@

```{r global-regional-tb-summary, fig.cap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017. Globally incidence rates have been declining since the early 2000s but this decline varies with region.", fig.scap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, viridis_end = 1) + 
  theme(text = element_text(size = 12)) +
  labs(y = "Estimated incidence (all forms) per 100,000 population")
```

@@

TB remains one of the top 10 causes death worldwide, leading to 1.7 million deaths in 2016 alone.[@TheWorldHealthOrganization:2018va] The absolute number of deaths due to TB has fallen since 2000, with an average annual global rate of decline in TB mortality rates of 2.9% between 2000-2016. Unlike the trend observed for incidence rates, the year-on-year decline of TB mortality rates has remained consistent in all regions (Figure \@ref(fig:global-regional-mortality-summary)). Several regions, including Africa and Europe, have seen TB mortality rates fall to below 50% of those in 2000.

@@

There is an ongoing global co-epidemic of HIV and TB, with people living with HIV accounting for 1.4 million TB cases in 2016. 22% of deaths from TB were in those living with HIV. Whilst this is a global problem, it is a particular issue in sub-Saharan Africa with over 60% (95% confidence interval (CI): 55%-64%) of incidence TB cases in South Africa also having HIV (Figure \@ref(fig:map-global-tb-with-hiv)). This compares to a global mean of 9.1% (95% CI: 6.0%-13.0%) and a mean of 26.7% (95% CI: 17.4%-38.1%) in Africa.

@@

Multi-drug-resistant TB (MDR-TB), which is defined as being resistant to at least isoniazid and rifampin, made up 4.6% of all incident TB cases globally in 2015 (480,000). It can be acquired both through treatment failure and through transmission. Treatment requires the use of second line antibiotics, which often have more severe side effects and are more likely to fail, with only 52% successfully treated globally compared to 83% for drug susceptible TB.[@WHOTB2016] As for HIV co-infection, drug resistance is globally heterogeneous with some regions, like countries in the former USSR, having a much higher proportion of drug resistant cases. Figure \@ref(fig:map-global-tb-with-rr) shows the country level proportion of cases with at least rifampicin resistance and highlights the higher level of rifampicin resistance in countries formerly in the USSR. 92% of rifampicin cases in Russia in 2018 also had MDR-TB, which is comparable to the global median of 81% (Table \@ref(tab:prop-rr-with-mdr-tab)). Across all regions, Europe had the highest median percentage of rifampicin cases with MDR-TB (87%), with the Western Pacific having the lowest (78%).

@@

```{r prop-rr-with-mdr-tab, warning = FALSE}
summarise_tb_burden(countries = "Russia", 
                    metric = "e_mdr_pct_rr_new", verbose = FALSE,
                    year = 2018, stat = "median") %>% 
  mutate_at(.vars = c("e_mdr_pct_rr_new", "e_mdr_pct_rr_new_lo", "e_mdr_pct_rr_new_hi"),
            funs(ifelse(. > 100, 100, .))) %>% 
  arrange(desc(area)) %>% 
  mutate(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)` = 
           pretty_ci(e_mdr_pct_rr_new,
                     e_mdr_pct_rr_new_lo, 
                     e_mdr_pct_rr_new_hi, digits = 1)) %>% 
  mutate(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)` = 
           ifelse(area == "Russian Federation", paste(pretty_round(e_mdr_pct_rr_new), "***"),
                  as.character(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)`)
                  )) %>% 
  select(`WHO region` = area, `Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)`) %>% 
  knitr::kable(
               caption = "Percentage (\\%) of rifampicin resistant TB cases that have multi-drug resistant TB in Russia and regional medians, with interquartile ranges. The estimate for Europe includes data from the Russian Federation.",
        booktabs = TRUE) %>% 
  add_footnote(label = c("** Multi-Drug Resistant TB", "*** Only point estimates available"), notation = "none") %>% 
  kable_styling(font_size = 8, latex_options = "HOLD_position") %>% 
  {.}
```

@@

The Bacillus Calmette–Guérin (BCG) vaccine was first given to humans in 1921 and remains the only licensed vaccine for TB.[@Medicine2013] The BCG vaccine is a live vaccine and was developed by weakening a strain of *Mycobacterium bovis*, which is commonly found in cows, over a period of 13 years.[@TheWorldHealthOrganization:2018va] Serious side effects are rare, although a small scar at the injection site is common. Initially, public acceptance was slow, with low take up until after the Second World War. However, controversy remains and several countries have scaled back, or retargeted, their usage in recent years. This section details the action, effectiveness, duration of protection, effects and usage of the BCG vaccine.

@@

The BCG primarily acts by directly preventing the development of active, symptomatic disease. However, there is some evidence to suggest that the BCG vaccine also provides partial protection against initial infection.[@Roy2014] There is no evidence that BCG vaccination of latent TB cases provides protection from developing active TB disease.[@TheWorldHealthOrganization:2018va]

@@

A meta-analysis of RCTs indicated that increased protection is associated with distance from the equator.[@Mangtani2014a] One hypothesis for this is that there is a greater density of mycobacteria near the equator that may mask, or block, the protection offered by the BCG vaccine.[@Zwerling2011] Recently it has been found that much of this latitude effect may be due to the robustness of tuberculin skin testing (TST), with lower stringency near the equator.[@TheWorldHealthOrganization:2018va] TST screening tests for the presence of TB infection but may give a false positive if the subject has been exposed to other mycobacteria or the BCG vaccine. Reduced stringency would lead to a greater number of latent TB individuals being vaccinated. These individuals would then receive no protection from the vaccine and would lead to a reduced estimate of the effectiveness of the vaccine overall. Regardless of the mechanism this effect means that vaccination early in life maximises the protection conferred by the BCG vaccine in countries with evidence of decreasing protection with age.

@@

In addition to its effect on TB outcomes, the BCG vaccine has also been found to be effective at preventing leprosy (with a risk ratio (RR_ of 0.45 (95% CI: 0.34-0.56)), with some evidence that this protection was stronger in those vaccinated before 15 years of age.[@TheWorldHealthOrganization:2018va] Additionally, there is some evidence that the BCG vaccine can provide protection against Non-TB mycobacteria (NTM) infections, with an estimated effectiveness of 50%.[@TheWorldHealthOrganization:2018va] This protection has been shown to greatly reduce the incidence of NTM lymph node inflammation and protect against Buruli ulcers for the first 12 months following vaccination (RR, 0.50 [95% CI, 0.37–0.69]). A study also found that individuals with Buruli ulcers are less likely to develop osteomyelitis (bone infection) if they have a BCG scar (RR, 0.36 (95% CI: 0.22 to 0.58)).[@Zimmermann:2018io]

@@

The BCG vaccine is one of the mostly widely-used vaccines worldwide, with approximately 100 million doses given annually.[@The2004] However, due to the variable estimates of BCG efficacy, vaccination has been controversial since its development. The WHO recommends vaccination for all neonates as early as possible after birth in high burden settings. Vaccination in low burden settings is dependent on the country specific epidemiology of TB.[@WHO2017; @TheWorldHealthOrganization:2018va] This recommendation is based on the strong evidence that the BCG is highly protective in children (particularly against disseminated TB and TB meningitis),[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014] Historically, different strategies have been utilized worldwide. These include universal vaccination of those at most risk of onwards transmission and high-risk group vaccination targeting either neonates or children.[@Abubakar2013]

@@

As of 2011, among 180 countries with available data, 157 countries recommended universal BCG vaccination. The remaining 23 countries had either never implemented a universal programme or have switched to targeted vaccination of high risk individuals.[@Zwerling2011a] Most countries began universal programmes between the 1940s and 1980s due to high levels of TB incidence and strong evidence of the effectiveness of the BCG vaccine.[@Hart1972] In the last 20 years 49 of these countries reported changing their vaccination programme with 27 countries reporting major changes in the last 10 years.[@Zwerling2011a] Globally, in countries that have BCG vaccination policies in place, coverage is thought to range from 70% to 100%.

### 2019-12-11 rmds/03-getTBinR.Rmd
dodd minor corrections up to 6

`getTBinR` is an R package[@RCoreTeam2019] that I developed to facilitate working with the data[@WHO:2018] collected by the World Health Organization (WHO) on the country level epidemiology of Tuberculosis (TB). All data is freely available from the WHO^[WHO TB data: https://www.who.int/tb/country/data/download/en/] and the package code is archived on Zenodo^[Zenodo: https://zenodo.org/badge/latestdoi/112591837] and Github^[GitHub: https://github.com/seabbs/getTBinR/]. The aim of `getTBinR` is to allow researchers, and other interested individuals, to quickly and easily gain access to a detailed TB dataset and to start using it to derive key insights. It provides a consistent set of tools that can be used to rapidly evaluate hypotheses on a widely used dataset before they are explored further using more complex methods or more detailed data. Prior to the development of `getTBinR` access to the WHO data was ad-hoc and there were no standardised visualisation or summary tools.

@@

The data sourced by `getBTinR` is collected by the WHO, via member governments, and used to compile the yearly global TB report.[@WHO:2018] Data collection encompasses TB incidence, TB mortality rates, the age distribution of TB cases, the proportion of drug resistant cases, case detection rates, and treatment rates. For a complete description of the data and data collection process, see [@WHO:2018]. These data are used by the WHO, governmental organisations and researchers to summarise country level TB epidemiology, as well as the wider global and regional picture.

@@

Or accessed online^[Dashboard: http://www.seabbs.co.uk/shiny/ExploreGlobalTB]. Any metric in the WHO TB data can be explored, with country selection using the built in map, and animation possible by year. The shiny app can also be used to generate the country level reports discussed in the next section. Figure \@ref(fig:dash-snapshot) shows a screenshot of the dashboard, with South Africa selected as the country of interest.

>>>>

`getTBinR` is an R package[@RCoreTeam2019] that I developed to facilitate working with the data[@WHO:2018] collated by the World Health Organization (WHO) on the country level epidemiology of Tuberculosis (TB). All data is freely available from the WHO^[WHO TB data: https://www.who.int/tb/country/data/download/en/] and the package code is archived on Zenodo^[Zenodo: https://zenodo.org/badge/latestdoi/112591837] and Github^[GitHub: https://github.com/seabbs/getTBinR/]. The aim of `getTBinR` is to allow researchers, and other interested individuals, to quickly and easily gain access to a detailed TB dataset and to start using it to derive key insights. It provides a consistent set of tools that can be used to rapidly evaluate hypotheses on a widely used dataset before they are explored further using more complex methods or more detailed data. Prior to the development of `getTBinR` access to the WHO data was ad-hoc and there were no standardised visualisation or summary tools.

@@

The data sourced by `getBTinR` is collated by the WHO, via member governments, and used to compile the yearly global TB report.[@WHO:2018] Data collation and estimation encompasses TB incidence, TB mortality rates, the age distribution of TB cases, the proportion of drug resistant cases, case detection rates, and treatment rates. For a complete description of the data and data collection process, see [@WHO:2018]. These data are used by the WHO, governmental organisations and researchers to summarise country level TB epidemiology, as well as the wider global and regional picture.

@@

Any metric in the WHO TB data can be explored, with country selection using the built in map, and animation possible by year. The shiny app can also be used to generate the country level reports discussed in the next section. Figure \@ref(fig:dash-snapshot) shows a screenshot of the dashboard, with South Africa selected as the country of interest.
